---
title: "ggtaichi"
output: github_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/PursuitOfDataScience/ggtaichi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PursuitOfDataScience/ggtaichi/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ggtaichi)](https://CRAN.R-project.org/package=ggtaichi)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.path = "man/figures/README-",
                      fig.align = "center",
                      dpi = 120)
```

`ggtaichi` is a `ggplot2` extension that compares data from two sources on a
single grid of taichi (yin-yang) diagrams. A regular heat map made with
`geom_tile()` encodes three dimensions (the `x`, `y` position and one value);
`geom_taichi()` turns every cell into a taichi symbol whose two interlocking
fish are filled by **two** sources at once, so four dimensions are expressed on
one plot.

## Installation

You can install the development version from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("PursuitOfDataScience/ggtaichi")
```

## Anatomy of a taichi

Each symbol is a circle split by an S-curve into two interlocking fish. The
**yang** (light) fish is shaded by one source and the **yin** (dark) fish by the
other, each on its own gradient. There are no decorative dots: every drop of ink
is data.

```{r anatomy, fig.width=4.5, fig.height=4.5, fig.alt="A single large taichi diagram whose red (yang) half encodes one data source and grey (yin) half the other."}
library(ggtaichi)
library(ggplot2)

one <- data.frame(x = 1, y = 1, google = 7, twitter = 3)

ggplot(one, aes(x, y)) +
  geom_taichi(yin = twitter, yang = google) +
  coord_fixed() +
  theme_taichi()
```

## A clear, small grid

The built-in `pitts_tg` dataset holds the 30-week COVID-related Google and
Twitter incidence rates for 9 categories in the Pittsburgh Metropolitan
Statistical Area. With many weeks the symbols shrink, so it is often easier to
read a slice. Here are the first six weeks, where each taichi is big enough to
compare the two halves at a glance.

```{r pitts-small, fig.width=7, fig.height=8, fig.alt="A six-week grid of taichi diagrams for nine Pittsburgh COVID categories, the yang fish shaded red by Google and the yin fish grey by Twitter."}
pitts_small <- subset(pitts_tg, week <= 6)

ggplot(pitts_small, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  theme_taichi() +
  ggtitle("Pittsburgh: Google (yang) vs Twitter (yin), weeks 1-6")
```

The legend titles default to the column names you supply. Note how `Covid` and
`Masks` lean dark (high Twitter) while staying pink (moderate Google).

## Your own palettes

Each fish gets its own gradient, and any extra argument is passed straight to
`ggplot2::scale_fill_gradientn()`.

```{r palettes, fig.width=7, fig.height=8, fig.alt="The same six-week Pittsburgh grid of taichi diagrams drawn with a blue gradient for Twitter and an orange gradient for Google."}
ggplot(pitts_small, aes(x = week, y = category)) +
  geom_taichi(
    yin  = Twitter, yin_name  = "Twitter (%)",
    yin_colors  = c("#deebf7", "#3182bd", "#08306b"),
    yang = Google,  yang_name = "Google (%)",
    yang_colors = c("#fee6ce", "#e6550d", "#7f2704")
  ) +
  theme_taichi()
```

## Comparing places

Because `geom_taichi()` is an ordinary layer, faceting just works. The
`states_tg` dataset repeats the same measurements across four states; showing two
of them over a handful of weeks keeps the glyphs large and legible.

```{r states, fig.width=7, fig.height=8, fig.alt="Two faceted taichi grids comparing New York and Texas over six weeks, the yang fish red for Google and the yin fish grey for Twitter."}
two_states <- subset(states_tg, state %in% c("New York", "Texas") & week <= 6)

ggplot(two_states, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  facet_wrap(~ state, ncol = 1) +
  remove_padding(x = "c", y = "d") +
  theme_taichi() +
  ggtitle("New York vs Texas, weeks 1-6")
```

See `vignette("ggtaichi")` for the full tour.

## Acknowledgement

`ggtaichi` is built on top of, and is the spiritual sibling of, the
[`ggDoubleHeat`](https://CRAN.R-project.org/package=ggDoubleHeat) package, which
introduced the idea of folding two data sources into a single reformed heat map
through the `geom_heat_*()` family. `ggtaichi` reuses that two-scale design (and
its example data) and re-imagines the per-cell glyph as a taichi diagram.
`ggDoubleHeat` is the foundational layer of this package and should be cited
when you use `ggtaichi`:

> Yu Y, Buskirk T (2025). *ggDoubleHeat: A Heatmap-Like Visualization Tool*.
> R package version 0.1.3. CRAN:
> <https://CRAN.R-project.org/package=ggDoubleHeat>, GitHub:
> <https://github.com/PursuitOfDataScience/ggDoubleHeat>

```
@Manual{,
  title  = {ggDoubleHeat: A Heatmap-Like Visualization Tool},
  author = {Youzhi Yu and Trent Buskirk},
  year   = {2025},
  note   = {R package version 0.1.3.
            GitHub: https://github.com/PursuitOfDataScience/ggDoubleHeat},
  url    = {https://CRAN.R-project.org/package=ggDoubleHeat},
}
```
