---
title: "Introduction to ggtaichi"
author: |
  Youzhi Yu
  <br><span style="font-size: 0.8em;">University of Chicago</span>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggtaichi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center"
)
```

## Why taichi?

A heat map drawn with `ggplot2::geom_tile()` carries three dimensions of
information: the `x` position, the `y` position, and a single value mapped to
fill. That is plenty when there is one number per cell, but it forces you to
*facet* (or to draw two separate maps) the moment you want to compare two data
sources on the same footing.

`ggtaichi` removes that limitation by replacing each cell with a **taichi**
(yin-yang) diagram. The symbol is a circle split by an S-curve into two
interlocking "fish":

- the **yang** (light) fish is shaded by one data source, and
- the **yin** (dark) fish is shaded by the other.

Because both fish live in the same cell, a single `geom_taichi()` layer encodes
**four** dimensions at once: `x`, `y`, `yin`, and `yang`. The two sources keep
their own color scales and legends, so they can be read independently while
still being compared side by side.

```{r}
library(ggtaichi)
library(ggplot2)
```

## The example data

`ggtaichi` ships with the same data sets used by its foundational package
`ggDoubleHeat`. `pitts_tg` records the 30-week COVID-related Google and Twitter
incidence rates for 9 categories in the Pittsburgh Metropolitan Statistical
Area (MSA).

```{r}
head(pitts_tg)
```

`states_tg` is the larger sibling, repeating the same measurements across four
states, and `pitts_emojis` holds the most popular weekly emoji per category.
See `?pitts_tg`, `?states_tg`, and `?pitts_emojis` for the full descriptions.

## A first taichi grid

The two value columns are passed to the `yin` and `yang` arguments. Everything
else — the `x`/`y` mapping, faceting, titles — is plain `ggplot2`.

```{r}
ggplot(pitts_tg, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  theme_taichi() +
  ggtitle("Pittsburgh Google & Twitter Incidence Rate (%)")
```

Each symbol stays round regardless of the panel's aspect ratio, so you do **not**
need `coord_fixed()`. The legend titles default to the column names you supplied
(`Twitter` and `Google` here).

## Reading a single symbol

It is worth zooming in on one cell to see the anatomy of the glyph. The yang
fish (its head at the bottom) carries one source; the yin fish (its head at the
top) carries the other. Each fish holds a small contrasting **eye**, exactly as
in a classic taichi symbol.

```{r, fig.width = 4.5, fig.height = 4.5}
one <- data.frame(x = 1, y = 1, a = 7, b = 3)

ggplot(one, aes(x, y)) +
  geom_taichi(yin = b, yang = a) +
  coord_fixed() +
  theme_taichi()
```

## Controlling the eyes

The eyes are decorative markers. Turn them off with `eyes = FALSE`, resize them
relative to the symbol with `eye_ratio`, and recolor them with `yin_eye` /
`yang_eye`.

```{r}
ggplot(pitts_tg, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google, eyes = FALSE) +
  theme_taichi() +
  ggtitle("Same grid, without the eyes")
```

## Customizing the color scales

Each fish gets its own gradient. `yang_colors` and `yin_colors` accept any color
vector (usually hex codes), and `yang_name` / `yin_name` relabel the legends.
Any extra argument is forwarded to `ggplot2::scale_fill_gradientn()`, so you can,
for example, set a common `limits` for both sources.

```{r}
ggplot(pitts_tg, aes(x = week, y = category)) +
  geom_taichi(
    yin = Twitter,  yin_name = "Twitter (%)",
    yin_colors = c("#deebf7", "#3182bd", "#08306b"),
    yang = Google, yang_name = "Google (%)",
    yang_colors = c("#fee6ce", "#e6550d", "#7f2704")
  ) +
  theme_taichi() +
  ggtitle("Custom palettes for each source")
```

## Removing the panel padding

`ggplot2` leaves a margin around discrete and continuous scales, which can make
a taichi grid look like it is floating. `remove_padding()` trims it; tell it
whether each axis is continuous (`"c"`) or discrete (`"d"`).

```{r}
ggplot(pitts_tg, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  remove_padding(x = "c", y = "d") +
  theme_taichi()
```

## Scaling up with `states_tg`

Because `geom_taichi()` is an ordinary layer, faceting works out of the box. The
`states_tg` data set carries the same measurements across four states.

```{r, fig.height = 9}
ggplot(states_tg, aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  facet_wrap(~ state, ncol = 1) +
  remove_padding(x = "c", y = "d") +
  theme_taichi() +
  ggtitle("Google & Twitter incidence rate by state")
```

## Acknowledgement

`ggtaichi` stands on the shoulders of the
[`ggDoubleHeat`](https://CRAN.R-project.org/package=ggDoubleHeat) package, which
pioneered the two-source "double" heat map through its `geom_heat_*()` family
and supplies the example data used throughout this vignette. Please cite it
alongside `ggtaichi`:

> Yu Y, Buskirk T (2025). *ggDoubleHeat: A Heatmap-Like Visualization Tool*.
> R package version 0.1.3. <https://CRAN.R-project.org/package=ggDoubleHeat>
