---
title: "Gallery"
author: |
  Youzhi Yu
  <br><span style="font-size: 0.8em;">University of Chicago</span>
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 6,
  fig.align = "center",
  dpi = 120
)
library(ggplot2)
library(ggtaichi)
```

A tour of what `geom_taichi()` can look like. Every plot below is a single
layer call plus ordinary ggplot2.

## One legend, two fish

When the two sources share units, `shared_legend = TRUE` paints both fish
with one palette on one scale, so the two halves of every glyph can be read
against a single legend. The bundled `cafes_tg` data (synthetic espresso vs.
matcha orders) is made for this:

```{r shared, fig.height = 6, fig.alt="A 12-week by 8-neighbourhood grid of taichi diagrams where espresso and matcha orders share one grey fill scale and a single legend."}
ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi(yin = matcha, yang = espresso,
              shared_legend = TRUE,
              yin_name = "orders / 100 customers") +
  remove_padding() +
  theme_taichi() +
  ggtitle("Espresso (yang) vs matcha (yin), one shared scale")
```

## Two palettes, shared limits

Keep each source's own palette but align the limits, so equal values carry
equal ink:

```{r shared-limits, fig.height = 6, fig.alt="The same espresso versus matcha grid with a blue palette for matcha and an orange palette for espresso on identical scale limits."}
ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi(yin = matcha,   yin_name = "Matcha",
              yin_colors  = c("#deebf7", "#3182bd", "#08306b"),
              yang = espresso, yang_name = "Espresso",
              yang_colors = c("#fee6ce", "#e6550d", "#7f2704"),
              shared_limits = TRUE) +
  remove_padding() +
  theme_taichi()
```

## Classic eyes, data-driven eyes

```{r eyes, fig.height = 4.2, fig.alt="A row of four taichi symbols whose white yin eyes grow and shrink with a data column while the black yang eyes stay a constant size."}
d <- data.frame(x = 1:4, y = 1, yin = c(2, 4, 6, 8), yang = c(8, 6, 4, 2),
                pull = c(30, 5, 18, 45))

ggplot(d, aes(x, y)) +
  geom_taichi(yin = yin, yang = yang, eyes = TRUE,
              yin_eye_size = pull, yang_eye_size = 0.12,
              limits = c(0, 10)) +
  coord_fixed() +
  theme_taichi() +
  ggtitle("Eye size as a fifth channel")
```

## A turning grid

Rotation can be pure annotation or a data channel — here each glyph's angle
encodes its column:

```{r rotation, fig.height = 6, fig.alt="A four-by-four grid of taichi diagrams whose rotation angle increases along the x direction."}
grid16 <- expand.grid(x = 1:4, y = 1:4)
grid16$yin <- seq(1, 10, length.out = 16)
grid16$yang <- rev(grid16$yin)
grid16$turn <- grid16$x * 22.5

ggplot(grid16, aes(x, y)) +
  geom_taichi(yin = yin, yang = yang, angle = turn, eyes = TRUE,
              limits = c(0, 10)) +
  coord_fixed() +
  theme_taichi()
```

## Categorical fills

```{r categorical, fig.height = 6, fig.alt="A three-by-three grid of taichi diagrams filled by discrete categories on both fish."}
d9 <- expand.grid(x = 1:3, y = 1:3)
d9$roast  <- factor(c("light", "medium", "dark")[(d9$x + d9$y) %% 3 + 1])
d9$origin <- factor(c("blend", "single")[(d9$x * d9$y) %% 2 + 1])

ggplot(d9, aes(x, y)) +
  geom_taichi(yin = roast, yang = origin) +
  coord_fixed() +
  theme_taichi()
```

## Texture at scale

Dense grids stop being symbols you read one by one and become a texture of
two interleaved fields — still useful for spotting bands and regime changes:

```{r texture, fig.height = 7, fig.alt="A dense 31-week by 9-category grid of small taichi diagrams for two states, read as an overall texture rather than glyph by glyph."}
ggplot(subset(states_tg, state %in% c("New York", "Texas")),
       aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google) +
  facet_wrap(~ state, ncol = 1) +
  remove_padding() +
  theme_taichi() +
  ggtitle("31 weeks as texture")
```

## Bring your own scales

`yin_scale` / `yang_scale` accept any fill scale, and the exported
`geom_yin_fish()` / `geom_yang_fish()` let you assemble everything by hand
(your scales, your `ggnewscale` stacking):

```{r custom, fig.height = 5, fig.alt="A row of taichi diagrams using viridis palettes supplied as custom scales."}
ggplot(d, aes(x, y)) +
  geom_taichi(yin = yin, yang = yang,
              yin_scale = scale_fill_viridis_c,
              yang_scale = scale_fill_viridis_c(name = "yang", option = "magma")) +
  coord_fixed() +
  theme_taichi()
```

## The gap, drawn three ways

`explicit` computes the relationship between the two sources and shows it as a
third channel. The eyes are the default --- subordinate to the fills, so the
two sources stay the story, and absent altogether where the sources agree:

```{r explicit-eye, fig.height = 6, fig.alt="The espresso versus matcha grid with eyes whose size grows with the gap between the two sources; cells where they agree have no eyes."}
ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi(yin = matcha, yang = espresso,
              shared_legend = TRUE, yin_name = "orders / 100 customers",
              explicit = "difference") +
  remove_padding() +
  theme_taichi() +
  ggtitle("Eye size = the gap")
```

Tilt is the most accurate of the four channels, because direction is read far
more precisely than shading. Upright means the two sources agree:

```{r explicit-angle, fig.height = 6, fig.alt="The espresso versus matcha grid where each glyph leans left or right in proportion to which source is larger."}
ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi(yin = matcha, yang = espresso,
              shared_legend = TRUE, yin_name = "orders / 100 customers",
              explicit = "difference", explicit_channel = "angle") +
  remove_padding() +
  theme_taichi() +
  ggtitle("Tilt = the gap")
```

Or drop the glyph entirely when the gap *is* the question:

```{r diff, fig.height = 5, fig.alt="A diverging heatmap of matcha minus espresso orders, red where espresso leads and blue where matcha does."}
ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi_diff(yin = matcha, yang = espresso) +
  remove_padding() +
  theme_taichi() +
  ggtitle("matcha - espresso")
```

## A fair pair of palettes

First the package defaults, then `palette = "balanced"`. The data is
symmetric --- both fish carry the same value in every cell --- so a fair pair
should make the two halves of every glyph look equally heavy. Only one of them
does:

```{r pair, fig.width = 9, fig.height = 3.6, fig.alt="Two rows of taichi diagrams drawn from identical data, the default grey and red palette above and a luminance-matched blue and brick-red pair below; only the matched pair makes the two halves look equally weighted."}
same <- data.frame(x = 1:6, y = 1, a = seq(1, 10, length.out = 6))
same$b <- same$a

both <- function(pal, title) {
  ggplot(same, aes(x, y)) +
    geom_taichi(yin = a, yang = b, palette = pal, shared_limits = TRUE,
                show.legend = FALSE) +
    coord_fixed() +
    theme_taichi() +
    ggtitle(title)
}

both("default", "palette = \"default\"")
both("balanced", "palette = \"balanced\"")
```

And the measurement behind it:

```{r check}
taichi_check_palette()
```

## Binned fills

On a grid too dense to compare cell by cell, matching a patch to one of a few
labelled bins beats reading a continuous ramp:

```{r binned, fig.height = 7, fig.alt="The dense two-state weekly grid with both fish filled from five discrete colour steps."}
ggplot(subset(states_tg, state %in% c("New York", "Texas")),
       aes(x = week, y = category)) +
  geom_taichi(yin = Twitter, yang = Google,
              yin_scale  = scale_taichi_yin_binned(n.breaks = 5),
              yang_scale = scale_taichi_yang_binned(n.breaks = 5),
              shared_limits = TRUE) +
  facet_wrap(~ state, ncol = 1) +
  remove_padding() +
  theme_taichi()
```

## Interactive: hover one source, highlight it everywhere

`interactive = TRUE` hands the layers to [ggiraph](https://davidgohel.github.io/ggiraph/). Hover a cell for the
exact values and their difference; with `data_id_by = "source"`, hovering any
yin fish highlights the yin fish in *every* cell, which turns the
superposition display into a single-source display for as long as you hold the
pointer there.

```{r girafe, eval = requireNamespace("ggiraph", quietly = TRUE), fig.height = 6}
p <- ggplot(cafes_tg, aes(x = week, y = neighbourhood)) +
  geom_taichi(yin = matcha, yang = espresso,
              shared_legend = TRUE, yin_name = "orders / 100 customers",
              interactive = TRUE, data_id_by = "source") +
  remove_padding() +
  theme_taichi()

ggiraph::girafe(
  ggobj = p,
  width_svg = 7, height_svg = 6,
  options = list(
    ggiraph::opts_hover(css = "stroke:#C20824;stroke-width:1.5px;"),
    ggiraph::opts_tooltip(
      css = "background:#f3efe6;border:1px solid #222;padding:6px;border-radius:3px;"
    )
  )
)
```
